\scuba: iterative map-making with the Sub-Millimetre User Reduction Facility

### 2012-09-02 smurf_paper.tex
some work on smurf algorithm section

\begin{figure*}
\centering
\includegraphics[width=\linewidth]{pca.pdf}
\caption{The first twelve components from a principal component
  analysis, ranked by the mean eigenvalues, of combined 450 (s4a
  subarray) and 850\,\micron\ (s8b subarray) time-series bolometer
  data, for the same observation as that used in the right-hand panels
  of Figs.~\ref{fig:bolos_mix} and \ref{fig:pspec}. For each
  component, the top plot shows the time series of the eigenvector,
  normalized by its \rms. The bottom coloured panels show the
  eigenvalues for the bolometers at each wavelength (the amplitude of
  the eigenvector in the time series). For reference, both the mean,
  $\bar{\lambda}$, and \rms, $\lambda_\mathrm{rms}$, eigenvalues for
  the bolometers in each subarray are also shown. The first two
  components have comparable amplitudes, and dominate by at least a
  factor of 10 any other components of the time series. We suspect
  that they are caused by atmospheric variations (expected to be
  brighter at 450\,\micron), and fridge oscillations (expected to be
  comparable at each wavelength). However, the two eigenvectors do not
  necessarily map directly to these physical sources under this
  interpretation; rather they are two \emph{different} linear
  combinations of the underlying physical signals that give rise to
  statistically uncorrelated components. The third component
  accurately reproduces an apparent data-acquisition glitch that
  appears only at 450\,\micron. The fourth component is a periodic
  feature only observed in column 15 of the 450\,\micron\
  subarray. The fifth component shows a series of steps and glitches
  that appear predominantly along the bottom-left edge of the
  850\,\micron\ subarray where the bolometers are poorly
  biased. Finally, the sixth component is degenerating into a less
  significant, and more randomly scattered signal across the focal
  plane. Subsequent components follow this trend, and would generally
  be kept in a PCA-cleaning procedure.}
\label{fig:pca}
\end{figure*}

@@

Components~6, and 8--10 in Fig.~\ref{fig:pca} have several generic
features in common. First, the PSDs exhibit moderate low-frequency
$1/f$ noise. Second, there are broad-line features at frequencies
above $\sim$30\,Hz which we believe to be aliased high-frequency
noise, in part due to the clear correlation patterns that appear along
columns in the eigenvalue maps at both wavelengths (e.g. columns 26
and 29 at 450\,\micron in Component~6). Finally, these components also
have potentially unrelated features mixed into them, like sudden
offsets that are likely residuals due to imperfect step correction
(e.g. $\sim$50\,s in the eigenvector time-series for Component~6; also
see Section~\ref{sec:steps}).

@@

@@

The remainder of this section gives an overview of the model that is
assumed by SMURF for the \scuba\ data, and the typical steps used to
estimate a map. Subsequent sections discuss problems with this model
and additional processing steps and constraints that can be used to
improve the solution in different cases.

@@

\item Describe the bolometer signal model, and with the help of a flow-chart
show the basic procedure that SMURF follows.

@@

We express the signal observed by the $i$th bolometer as a function of
 time, \textbf{Per: $e_i(t)$ is written as a vector ( bold ) but I assume
 we use it as a scalar.}

\begin{equation}
\mathbf{b}_i(t) = g_i[\mathbf{e}_i(t) \mathbf{a}_i(t) + \mathbf{n}_i(t)],
\end{equation}
where $\mathbf{a}(t)$ is the time-varying signal produced by scanning
the telescope across astronomical sources, $\mathbf{e}(t)$ is the
time-varying extinction, which is a function of the telescope
elevation and atmospheric conditions, and $\mathbf{n}_i(t)$ represents
sources of noise. The two terms in square brackets, as written, have
units of power delivered to the detectors (pW). The scale factor $g_i$
converts this effective power to the digitized units recorded on disk
(DAC) -- the \emph{flatfield} multiplied by a digitization constant --
and in this formulation it is assumed to be constant in time.

@@

%-------------------------------------------------
\subsection{Data preparation}
%-------------------------------------------------

>>>>

\begin{figure*}
\centering
\includegraphics[width=\linewidth]{pca.pdf}
\caption{The first twelve components from a principal component
  analysis, ranked by the mean eigenvalues, of combined 450 (s4a
  subarray) and 850\,\micron\ (s8b subarray) time-series bolometer
  data, for the same observation as that used in the right-hand panels
  of Figs.~\ref{fig:bolos_mix} and \ref{fig:pspec}. For each component
  the top plot shows the time series of the normalized eigenvectors,
  the middle plot its PSD, and the bottom coloured panels the
  eigenvalues for the bolometers at each wavelength (the amplitudes of
  the eigenvectors in the time series). For reference, both the mean,
  $\bar{\lambda}$, and \rms, $\lambda_\mathrm{rms}$, eigenvalues for
  the bolometers in each subarray are also shown. Most of the
  correlated $1/f$ signal is encapsulated in Components~1 and 2, and
  appears to be dominated by scan-synchronous noise (compare with the
  scan pattern in the right-hand panel of Fig.~\ref{fig:bolos_mix}).
  Components~3, 4, 7 and 11 are transient glitches that only appear in
  a small subset of the bolometers. Component~5 also appears in only a
  few bolometers, and is dominated by a narrow line at 57\,Hz which
  suggests that it is related to the 60\,Hz mains. Components~6 and
  8--10 contain a mixture of broad lines above $\sim$30\,Hz (which are
  probably aliased high-frequency noise), with some low-frequency
  drifts. Finally, the erratic appearance of the eigenvector for
  Component~12 is typical of a poorly-biased detector.}
\label{fig:pca}
\end{figure*}

@@

Components~6 and 8--10 in Fig.~\ref{fig:pca} have several generic
features in common. First, the PSDs exhibit moderate low-frequency
$1/f$ noise. Second, there are broad-line features at frequencies
above $\sim$30\,Hz which we believe to be aliased high-frequency
noise, in part due to the clear correlation patterns that appear along
columns in the eigenvalue maps at both wavelengths (e.g. columns 26
and 29 at 450\,\micron in Component~6). Finally, these components also
have potentially unrelated features mixed into them, like sudden
offsets that are likely residuals due to imperfect step correction
(e.g. $\sim$50\,s in the eigenvector time-series for Component~6; also
see Section~\ref{sec:steps}).

@@

This example illustrates some of the types of correlated signals and
patterns that PCA can identify in \scuba\ data. While there are
typically one or two strong signal components detected, in general,
the details can vary significantly from data set to data set, and the
lengths of the time-series analyzed. The reason for this is probably
due to the fact that many of the noise sources are not stationary in
time (e.g. scan-synchronous noise which obviously depends on the scan
pattern). It should also be clear from this example that while PCA
offers some helpful insight into the various sources of noise, it does
not necessarily identify clean patterns that can easily be
modeled. For example, while there are clearly correlation patterns
along columns (Components~6 and 8--10, particularly at 450\,\micron),
as one might expect given the common amplification chain, the
intensities of these high-frequency signals are random in the
eigenvalue maps, and a common-mode signal estimated from the data for
each column would not remove it.

@@

In practice, the model generally adopted for low-frequency noise
\scuba\ data is a single common-mode (average signal across all
working bolometers in a given focal plane), with most remaining noise
sources being removed by a high-pass filter. Many variations on this
were attempted throughout the commissioning phase, but as the PCA in
Secion~\ref{sec:pca} illustrates, less significant noise sources are
both transient and particularly difficult to model.

This section gives an overview of the SMURF algorithm, including data
cleaning and other pre-processing steps (Section~\ref{sec:dataprep}),
the iterative estimation of the map and noise model components
(Section~\ref{sec:components}), and concludes with a discussion of
convergence (Section~\ref{sec:converge}).

@@

@@

We express the signal observed by the $i$th bolometer as a function of
time,
%
\begin{equation}
\mathbf{b}_i(t) = g_i[\mathbf{e}_i(t) \mathbf{a}_i(t) + \mathbf{n}_i(t)],
\end{equation}
%
where $\mathbf{a}(t)$ is the time-varying signal produced by scanning
the telescope across astronomical sources, $\mathbf{e}(t)$ is the
time-varying extinction, which is a function of the telescope
elevation and atmospheric conditions, and $\mathbf{n}_i(t)$ represents
sources of noise. The two terms in square brackets, as written, have
units of power delivered to the detectors (pW). The scale factor $g_i$
converts this effective power to the digitized units recorded on disk
(DAC) -- the \emph{flatfield} multiplied by a digitization constant --
and in this formulation it is assumed to be constant in time.

@@

%-------------------------------------------------
\subsection{Data preparation}
\label{sec:dataprep}
%-------------------------------------------------
